preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-07 Manuscript/manuscript.Rmd
Concluding paragraph polishing

Here we present *preciseTAD*, an R package for precise identification of TAD(or loop) boundaries under a supervised machine learning (ML) framework. Our method builds upon an optimized random forest (RF) classification model that leverages high resolution ChIP-seq data in the form of regulatory elements (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS). We introduce a novel technique for building RF classifiers in a boundary prediction setting composed of a resolution-specific binning process and feature engineering procedure. *preciseTAD* then employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select the most biologically meaningful boundary coordinates on the linear genome, thereby circumventing resolution restrictions of conventional TAD- and chromatin loop-callers. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, highlighting their biological significance. This improved precision in TAD and loop boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome. [??? Rewrite. Say the main methodology first - model on 4 TFBSs, uses distance, RUS, technique for bp-level prediction, one sentence per point. Detail next - BroadHMM and histone were tested, OC, OP were tested, various sampling techniques. Conclude with findings and package availability. For MD, last part when ready to write an abstract]

>>>>

Here we present *preciseTAD*, an R package for precise identification of TAD(or loop) boundaries under a supervised machine learning (ML) framework. Our method builds upon an optimized random forest (RF) classification model that leverages high resolution genome annotation data in the form of chromatin segmentation (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS). We introduce a novel pipeline for building optimal boundary prediction classifier composed of a resolution-specific binning process and feature engineering procedure. We found that spatial associations between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level, *preciseTAD* employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select the most biologically meaningful boundary coordinates, thereby circumventing resolution restrictions of conventional TAD- and chromatin loop-callers. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, highlighting their biological significance. This improved precision in TAD and loop boundary location can provide insight into the role that epigenomics has on the 3D architecture of the human genome.
